---
title: "Stable Isotope Mixing Models with Covariates in R using cosimmr"
author: "Emma Govan and Andrew Parnell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette:
  toc: true
vignette: >
  %\VignetteIndexEntry{cosimmr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


## Introduction

`cosimmr` is a package in R designed for running Stable Isotope Mixing Models in
R, using Fixed Form Variational Bayes. The aim of this vignette is to allow users
to quickly learn how to use the package and giving examples of all the 
features within the package.
If you find any bugs in the software, or have new features which you think would be 
useful, please add this to the [Github issues page](https://github.com/emmagovan/cosimmr/issues)

## Installation of the `cosimmr` package

First, start Rstudio and find the window with the command prompt (the symbol `>`). Type

```{r,eval = FALSE}
install.packages("cosimmr")
```
It may ask you to pick your nearest CRAN mirror (the nearest site which hosts R packages). You will then see some activity on the screen as the `cosimmr` package and the other packages it uses are downloaded. The final line should then read: 

`package 'cosimmr' successfully unpacked and MD5 sums checked`

You then need to load the package. Type

```{r,eval = FALSE}
library(cosimmr)
```

This will load the `cosimmr` package and all the associated packages. You’ll need to type the `library(cosimmr)` command every time you start R.


There is some sample data sets (from ADD PAPER REF) available within cosimmr. 
Use the following command to access one.
```{r}
data(geese_data_day1)
```

This data can then be loaded into cosimmr using the function `cosimmr_load`

```{r}
cosimmr_1 <- with(
  geese_data_day1,
  cosimmr_load(
    formula = mixtures ~ 1,
    source_names = source_names,
    source_means = source_means,
    source_sds = source_sds,
    correction_means = correction_means,
    correction_sds = correction_sds,
    concentration_means = concentration_means
  )
)
```
This is a simple example that doesn't include any covariates. Note the formula is given in the form `tracer_data ~ covariates`. 

An isospace plot can be generated using this data
```{r}
plot(cosimmr_1)
```

The data can then be run through `cosimmr_ffvb`, the main function of the `cosimmr` 
package.
```{r}
cosimmr_1_out = cosimmr_ffvb(cosimmr_1)
```

Summary statistics for the run can then be viewed, several options are available, such as "statistics" and "quantiles"
```{r}
summary(cosimmr_1_out, type = "statistics")
```

The output of this can be plotted, using the plot function. There are several different options available in the plot function.
```{r}
plot(cosimmr_1_out, type ="beta_histogram", ind = 7)
```

More complex examples can also be run. Here we include the group the geese belong to as a covariate
```{r}
data(geese_data)
Groups = matrix(geese_data$groups)

cosimmr_2 <-cosimmr_load(
    formula = geese_data$mixtures ~ as.factor(Groups) -1,
    source_names = geese_data$source_names,
    source_means = geese_data$source_means,
    source_sds = geese_data$source_sds,
    correction_means = geese_data$correction_means,
    correction_sds = geese_data$correction_sds,
    concentration_means = geese_data$concentration_means)

cosimmr_2_out = cosimmr_ffvb(cosimmr_2)
```

Again we can view summary statistics for this run
```{r}
summary(cosimmr_2_out, type = "statistics")
```


We can then use the `predict` function to predict for each group
```{r}
pred_matrix = data.frame(Groups = c("Period 1", "Period 3"))
groups_predict = predict(cosimmr_2_out, x_pred = pred_matrix)
```
