---
title: "Stable Isotope Mixing Models with Covariates in R using cosimmr"
author: "Emma Govan and Andrew Parnell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette:
  toc: true
vignette: >
  %\VignetteIndexEntry{cosimmr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


## Introduction

`cosimmr` is a package in R designed for running Stable Isotope Mixing Models in
R, using Fixed Form Variational Bayes. The aim of this vignette is to allow users
to quickly learn how to use the package and giving examples of all the 
features within the package.
If you find any bugs in the software, or have new features which you think would be 
useful, please add this to the [Github issues page](https://github.com/emmagovan/cosimmr/issues)

## Installation of the `cosimmr` package

First, start Rstudio and find the window with the command prompt (the symbol `>`). Type

```{r,eval = FALSE}
install.packages("cosimmr")
```
It may ask you to pick your nearest CRAN mirror (the nearest site which hosts R packages). You will then see some activity on the screen as the `cosimmr` package and the other packages it uses are downloaded. The final line should then read: 

`package 'cosimmr' successfully unpacked and MD5 sums checked`

You then need to load the package. Type

```{r,eval = FALSE}
library(cosimmr)
```

This will load the `cosimmr` package and all the associated packages. You’ll need to type the `library(cosimmr)` command every time you start R.


There is some sample data sets (from ADD PAPER REF) available within cosimmr. 
Use the following command to access one.
```{r}
data(geese_data_day1)
```

This data can then be loaded into cosimmr using the function `cosimmr_load`

```{r}
group = c(rep("Period 1", 9))
cosimmr_1 <- with(
  geese_data_day1,
  cosimmr_load(
    formula = mixtures ~ 1,
    source_names = source_names,
    source_means = source_means,
    source_sds = source_sds,
    correction_means = correction_means,
    correction_sds = correction_sds,
    concentration_means = concentration_means
  )
)

cosimmr_1 <- cosimmr_load(
    formula = geese_data_day1$mixtures ~ c(1,1,1,1,1,1,1,1,1.1),
    source_names = geese_data_day1$source_names,
    source_means = geese_data_day1$source_means,
    source_sds = geese_data_day1$source_sds,
    correction_means = geese_data_day1$correction_means,
    correction_sds = geese_data_day1$correction_sds,
    concentration_means = geese_data_day1$concentration_means
  )




```

An isospace plot can be generated using this data

```{r}
plot(cosimmr_1)
```

The data can then be run through `cosimmr_ffvb`, the main function of the `cosimmr` 
package.
```{r}
cosimmr_1_out = cosimmr_ffvb(cosimmr_1)

summary(cosimmr_1_out)
```


The output of this can be plotted, using the plot function. There are several different
options available in the plot function
```{r}
plot(cosimmr_1_out, type = c("isospace", "beta_histogram", "p_ind"), ind = c(1,7))
```

Predictions
```{r}
cosimmr_predict(cosimmr_1_out, c(1,3))

cosimmr_predict(cosimmr_1_out, c(1))

newdata = matrix(c(1,3), ncol = 1)

a<-cosimmr_predict(cosimmr_1_out, matrix(c(1,3), ncol = 1))

cosimmr_predict(cosimmr_1_out, as.matrix(1))
```







Categorical variables also work
```{r}
colour_cat = c("red", "green", "blue", "red", "red", "red", "blue", "green", "blue")

# data(geese_data)
# cosimmr_2 <- with(
#   geese_data,
#   cosimmr_load(
#     formula = mixtures ~ groups,
#     source_names = source_names,
#     source_means = source_means,
#     source_sds = source_sds,
#     correction_means = correction_means,
#     correction_sds = correction_sds,
#     concentration_means = concentration_means
#   )
# )
data(geese_data)

Groups =matrix(geese_data$groups)
Groups = matrix(c(rep(1,9), rep(2,29), rep(3,74), rep(4,10), 
               rep(5,41), rep(6,20), rep(7,32), rep(8,36)), ncol = 1)

colnames(Groups) = c("Groups")


cosimmr_2 <-cosimmr_load(
    formula = geese_data$mixtures ~ as.factor(Groups),
    source_names = geese_data$source_names,
    source_means = geese_data$source_means,
    source_sds = geese_data$source_sds,
    correction_means = geese_data$correction_means,
    correction_sds = geese_data$correction_sds,
    concentration_means = geese_data$concentration_means)

cosimmr_2_out = cosimmr_ffvb(cosimmr_2)

summary(cosimmr_2_out, individual = c(1, 10, 39, 113,123,164,184, 251))

df = data.frame(Groups = c(1,2,3))


pre<-predict(cosimmr_2_out, df)


 round(colMeans(pre[1,,]),3)
 sd = c(rep(NA, 4))
 
 for(i in 1:4){
   sd[i] = round(sd(pre[1,,i]),3)
 }
 
 print(sd)

```




#run by 'hand' using rcpp code
```{r}
########## SET THESE
data(geese_data)
Groups <- geese_data$groups


x <- matrix(c(Groups), ncol = 1)
x_scaled <- cbind(matrix(rep(1, 251), ncol = 1), x)
#x_scaled <- matrix(c(rep(1, n)), ncol = 1)

#x_scaled = cbind(matrix(rep(1,251)), model.matrix(geese_data$mixtures ~ Groups))
x_scaled = model.matrix(geese_data$mixtures ~ Groups)
n_covariates <- (ncol(x_scaled))
y <- geese_data$mixtures

#-------------------- FFVB variables-----------------------
#Don't actually need to edit anything here, just run it


# Variables for the FFVB
S <- 100
K = 4
mu_beta_zero <- matrix(c(rep(0, K * (n_covariates))), 
                       nrow = (n_covariates), 
                       ncol = K)
#n_covariates + 1 for alpha (or beta_0) and then 1 beta for each covariate
sigma_beta_zero <- matrix(c(rep(1, K * (n_covariates))), 
                          nrow = (n_covariates), 
                          ncol = K)


mu_c = geese_data$correction_means
mu_s = geese_data$source_means
q = geese_data$concentration_means
sigma_c = geese_data$correction_sds
sigma_s = geese_data$source_sds
n_isotopes <- ncol(mu_c)
c_0 <- c(rep(1, n_isotopes)) #Change to 0.0001
d_0 <- c(rep(1, n_isotopes))
beta_lambda<-c(rep(0, K),rep(1, K * (K + 1) / 2))
lambda <- c(
  rep(beta_lambda, n_covariates),
  rep(1, n_isotopes), #shape
  rep(1, n_isotopes) #rate
)

c_prior = c(rep(1, 251))

lambdaout <- run_VB_cpp(lambda, K, n_isotopes, n_covariates, 251, 1, 
                        as.matrix(q), as.matrix(mu_s), as.matrix(mu_c), 
                        as.matrix(sigma_c), as.matrix(sigma_s), y, x_scaled,
                        100, 10, 0.9, 0.9, 100, 0.1, 50, c_prior, 
                        solo = FALSE)


function (lambdastart, n_sources, n_tracers, n_covariates, n, 
  beta_prior, concentrationmeans, sourcemeans, correctionmeans, 
  corrsds, sourcesds, y, x_scaled, S, P, beta_1, beta_2, tau, 
  eps_0, t_W, c_prior, solo) 
  
n_samples = 3600

theta_out_rcpp <- sim_thetacpp(S, lambdaout, K, n_isotopes, n_covariates)
```










